Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-11 paper/manuscript.Rmd
conclusion

For several decades after their introduction to Fogo Island, the caribou
population did not exceed ~100 individuals [@Bergerud1989, Newfoundland and
Labrador Wildlife Division, unpublished data]. During the 1990s, population
density reached ~300 animals and it is unknown when caribou began to occupy other
islands, although anecdotal evidence suggests it was during the 1990s. We surmise that as
competition among conspecifics increased along with population size,
density-dependent habitat selection resulted in expansion of the population to
nearby islands. Following Ideal Free Distribution theory, caribou should swim to
new islands when the average fitness of individuals on the starting island
exceeds the density-fitness equilibrium [@Morris1987]. While this is an ultimate
explanation, our data suggest caribou only remain on smaller islands for a few
days at a time, so the fitness equilibrium remains a theoretical construct. More
likely, however, is that fine-scale competition, density-dependent habitat
selection, and forage depletion drives individuals to periodically swim
between islands. Given the presumed low density of coyotes on Fogo Island and
the minimal risk they pose, predation risk does not appear to be a major driver
of swimming.

Our observations add to the evidence that caribou can, and occasionally do, swim
in the ocean [e.g. @Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. 
We suggest that swimming is likely more common than previously thought
for caribou living on oceanic islands, and that forage limitation and the
associated density-dependent habitat selection is an ultimate explanation for
this phenomenon. Although our inference is limited to observations, islands
appear to represent discrete foraging patches for terrestrial animals that can
influence fitness via increased foraging opportunities. We also posit that
individuals can vary in their evaluation of costs and benefits of movement
between these habitat patches.

>>>>

We surmise that as competition among conspecifics increased along with
population size, density-dependent habitat selection resulted in expansion of
the population to nearby islands. Following Ideal Free Distribution theory,
caribou should swim to new islands when the average fitness of individuals on
the starting island exceeds the density-fitness equilibrium [@Morris1987]. While
this is an ultimate explanation, our data suggest caribou only remain on smaller
islands for a few days at a time, so the fitness equilibrium remains a
theoretical construct. More likely, however, is that fine-scale competition,
density-dependent habitat selection, and forage depletion drives individuals to
periodically swim between islands. Given the presumed low density of coyotes on
Fogo Island and the minimal risk they pose, predation risk does not appear to be
a major driver of swimming. Our observations add to the evidence that caribou
can, and occasionally do, swim in the ocean [e.g. @Miller1995; @Miller2002;
@Jeffery2007; @Ricca2012]. We suggest that swimming is likely more common than
previously thought for caribou living on oceanic islands, and that forage
limitation and the associated density-dependent habitat selection is an ultimate
explanation for this phenomenon. Although our inference is limited to
observations, islands appear to represent discrete foraging patches for
terrestrial animals that can influence fitness via increased foraging
opportunities. We also present several hypotheses related to the nuances and
further variation within the idea that density-dependent habitat selection
governs swimming behaviour, and posit that above and beyond these additional
factors, individuals can vary in their evaluation of costs and benefits of
movement between these habitat patches.
